Definition of Archipelagos

### 2025-01-27 application_examples/abc_inferences/archipelago-definition.Rmd
Updating Rmd file for archipelago definition

---
title: "Definition of Archipelagos"
author: "Sara"
date: "`r Sys.Date()`"
output: html_document
bibliography: references.bib
---

@@

```{r, warning=FALSE, message=FALSE}
# If missing, install required packages

@@

```{r}
dpp4_chr <- "NC_036801.1"
dpp4_start <- 29502362
dpp4_end <- 29517382
region_start <- dpp4_start - 5e5
region_end <- dpp4_end + 5e5
```

@@

```{r}
# Load the GFF file
gff_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/ncbi_dataset/ncbi_dataset/data/GCF_000238955.4/genomic.gff" 
gff <- as.data.frame(import(gff_file))

@@

```{r}
table(gff$source)
```

@@

### Filter region of interest

```{r}
region_gff <- gff[
  gff$seqnames == "NC_036801.1" & 
  gff$type == "gene" & 
  gff$start <= region_end & 
  gff$end >= region_start, 
]

head(region_gff)
```

```{r}
table(region_gff$source)
```

```{r}
region_gff[grep("GeneID:101483164", region_gff$Dbxref), ]
```

```{r}
length(unique(region_gff$ID))
unique(region_gff$ID)
```

```{r}
# Extract relevant columns
gene_data <- region_gff[, c("strand","start", "end", "gene")]

@@

```{r}
# Set up larger plotting area and adjust margins
par(mar=c(6, 10, 4, 2), xpd=TRUE)

@@

# Add a legend at the top-left corner
legend("topleft", 
       legend=c("Positive strand (blue)", "Negative strand (red)"), 
       fill=c("blue", "red"), 
       bty="n", cex=0.8)
```

```{r}
# Sort gene_data by strand and then by start position
gene_data_sorted <- gene_data[order(gene_data$strand, gene_data$start), ]

# Initialize a vector to store overlap results
gene_data_sorted$overlap <- FALSE

# Loop over genes, comparing each gene with the next one in the same strand
for (i in 1:(nrow(gene_data_sorted) - 1)) {
  # Check if genes are on the same strand and overlap
  if (gene_data_sorted$strand[i] == gene_data_sorted$strand[i + 1] &&
      gene_data_sorted$end[i] >= gene_data_sorted$start[i + 1]) {
    gene_data_sorted$overlap[i] <- TRUE
    gene_data_sorted$overlap[i + 1] <- TRUE
  }
}

# View the genes with overlap information
head(gene_data_sorted)

```{r}
gene_data_sorted[which(gene_data_sorted$overlap==TRUE),"gene"]
```

```{r}
region_gff[region_gff$gene == "nop58",]
region_gff[region_gff$gene == "LOC111501074",]
```

The second one corresponds to the source `cmsearch`, so it may be the one that should be taken out.

```{r}
region_gff <- region_gff[region_gff$gene != "LOC111501074",]
```

### Define the 200 kb upstream regions

Calculate the upstream start and end to prevent overlap with the previous gene. *This is done for each strand from the second gene onwards as we dont know where the gene before the first gene is as it is not included in the region dataframe*.

- For genes on the positive strand, the upstream region starts 20kb before the gene's start position, or at the end of the previous gene if there is one.

- For genes on the negative strand, the upstream region starts 20kb after the gene's end position, or at the start of the next gene if there is one.

```{r}
# Separate genes by strand
positive_strand_genes <- region_gff[region_gff$strand == "+", ]
negative_strand_genes <- region_gff[region_gff$strand == "-", ]

# Sort genes by start position within each strand
positive_strand_genes <- positive_strand_genes[order(positive_strand_genes$start), ]
negative_strand_genes <- negative_strand_genes[order(negative_strand_genes$end, decreasing = TRUE), ]

# Initialize the new columns for upstream start and end
positive_strand_genes$upstream_start <- NA
positive_strand_genes$upstream_end <- NA

negative_strand_genes$upstream_start <- NA
negative_strand_genes$upstream_end <- NA

# Calculate the upstream region for positive strand genes
for (i in 2:nrow(positive_strand_genes)) {
  positive_strand_genes$upstream_start[i] <- max(positive_strand_genes$start[i] - 20000, positive_strand_genes$end[i-1] + 1)
  positive_strand_genes$upstream_end[i] <- positive_strand_genes$start[i] - 1
}

# Calculate the upstream region for negative strand genes
for (i in 2:nrow(negative_strand_genes)) {
  negative_strand_genes$upstream_start[i] <- min(negative_strand_genes$end[i] + 20000, negative_strand_genes$start[i-1] - 1)
  negative_strand_genes$upstream_end[i] <- negative_strand_genes$end[i] + 1
}

# Combine the results back
region_gff <- rbind(positive_strand_genes, negative_strand_genes)

# View the updated dataframe with upstream regions
head(region_gff)

#### Chromosome lengths

In case the upstream regions were to be defined for all chromosomes/scaffolds, the chromosome lengths would be needed to be the maximum for the start of the upstream region of the genes on the negative strand.

```{r, eval = FALSE}
gbff_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/ncbi_dataset/ncbi_dataset/data/GCF_000238955.4/genomic.gbff" 
file_lines <- readLines(gbff_file)
# Extract the LOCUS line
locus_lines <- file_lines[grep("^LOCUS", file_lines)]
head(locus_lines)
```

```{r}
length(unique(gff$seqnames))
# Remove version suffix from seqnames
length(unique(sub("\\.\\d+$", "", gff$seqnames)))
```

It does not have several versions.

```{r, eval = FALSE}
# Extract chromosome names and lengths
chrom_info <- gsub("^LOCUS\\s+(\\S+)\\s+(\\d+)\\s+.*$", "\\1,\\2", locus_lines)
chrom_info <- strsplit(chrom_info, ",")
chrom_names <- sapply(chrom_info, `[`, 1)
chrom_lengths <- as.numeric(sapply(chrom_info, `[`, 2))
# Create a named vector
chrom_lengths <- setNames(chrom_lengths, chrom_names)

# Remove version numbers from seqnames in the region-filtered gff
region_gff$seqnames <- sub("\\.\\d+$", "", region_gff$seqnames)

@@

```{r}
# Load the CGI annotation file
cgi_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/CGI-Mzebra.txt" 
cgif <- read.table(cgi_file, header = TRUE, sep = "\t")
head(cgif)
```

```{r}
region_cgi <- cgif[cgif$chr==dpp4_chr,]
```

```{r}
# Create a matrix of CGI intervals for faster lookup
cgi_intervals <- cbind(region_cgi$start, region_cgi$end)

# Function to efficiently check if positions are within any CGI
is_within_cgi_vector <- function(positions, cgi_intervals) {
  # Apply the check for each position against all intervals
  sapply(positions, function(pos) {
    any(pos >= cgi_intervals[,1] & pos <= cgi_intervals[,2])
  })
}

# Initialize results list for better memory management
results_list <- vector("list", nrow(region_gff))

# Process each gene
for (i in 1:nrow(region_gff)) {
  upstream_start <- region_gff$upstream_start[i]
  upstream_end <- region_gff$upstream_end[i]

  # Skip if NA values
  if (is.na(upstream_start) | is.na(upstream_end)) {
    next
  }

  # Create sequence of positions (handling both forward and reverse cases)
  if (upstream_start > upstream_end) {
    positions <- seq(from = upstream_start, to = upstream_end, by = -1)
  } else {
    positions <- seq(from = upstream_start, to = upstream_end, by = 1)
  }

  # Get CGI status for all positions at once
  in_cgi <- is_within_cgi_vector(positions, cgi_intervals)

  # Calculate distances properly based on strand
  distances <- if(region_gff$strand[i] == "+") {
    # For positive strand: distance from gene start
    region_gff$start[i] - positions  # Positive values will be upstream
  } else {
    # For negative strand: distance from gene end
    positions - region_gff$end[i]    # Positive values will be upstream
  }

  # Create result data frame for this gene
  results_list[[i]] <- data.frame(
    gene_id = region_gff$gene[i],
    position = positions,
    in_cgi = in_cgi,
    distance = distances,
    strand = region_gff$strand[i]
  )
}

# Combine all results efficiently
upstream_data <- do.call(rbind, results_list[!sapply(results_list, is.null)])

# Calculate counts and probabilities
position_counts <- table(upstream_data$distance)  # Using table() for correct counts
distance_seq <- as.numeric(names(position_counts))

# Calculate probabilities
cgi_probabilities <- tapply(upstream_data$in_cgi, 
                          upstream_data$distance, 
                          mean)   # Probability of being in CGI

```{r, fig.height=12, fig.width=10}
# Set up the plot with better formatting
par(mfrow = c(2,1))  # Set up 2 plots vertically
par(mar = c(4, 5, 4, 2))  # Adjust margins for top plot
# Plot 1: Probability plot
plot(distance_seq, cgi_probabilities, 
     xlab = "",  # Remove x-label from top plot
     ylab = "Probability of Position Being in CGI",
     main = "Probability of Position Being in CpG Island\nby Distance to Gene Start",
     cex.lab = 1.2,      # Larger axis labels
     cex.axis = 1.1,     # Larger axis numbers
     cex.main = 1.3,     # Larger title
     las = 1,            # Horizontal axis labels
     pch = 16,          # Solid circle points
     ylim = c(0, 1),    # Force probability y-axis to [0,1]
     col = adjustcolor("darkblue", alpha.f = 0.5))  # Semi-transparent blue

# Plot 2: Count plot
par(mar = c(5, 5, 3, 2))  # Adjust margins for bottom plot
plot(distance_seq, position_counts, type = "h",  # Using the correct counts
     xlab = "Distance to Gene Start (bp)",
     ylab = "Number of Positions",
     main = "Number of Positions at Each Distance",
     col = "darkgreen",
     lwd = 1,
     ylim = c(0, max(position_counts) * 1.05))  # Force y-axis to start at 0 with 5% padding
grid(col = "gray90", lty = "dotted")
```

@@

>>>>

---
title: "Definition of Archipelagos"
author: "Sara"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
bibliography: references.bib
---

@@

```{r load libraries, warning=FALSE, message=FALSE, echo=FALSE}
# If missing, install required packages

@@

```{r set region values}
dpp4_chr <- "NC_036801.1"
dpp4_start <- 29502362
dpp4_end <- 29517382
region_start <- dpp4_start - 5e5
region_end <- dpp4_end + 5e5
```

@@

```{r load gff}
# Load the GFF file
gff_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/ncbi_dataset/ncbi_dataset/data/GCF_000238955.4/genomic.gff" 
gff <- as.data.frame(import(gff_file))
names(gff)
```

@@

Lets keep the relevant columns:

```{r gff filter relevant columns}
gff <- gff[,c("seqnames", "start", "end", "width", "strand", "type", "source", "ID", "Dbxref", "gene")]
```

```{r gff table of annotation type}
table(gff$type)
```

```{r gff table of annotation source}
table(gff$source)
```

@@

### Filter genes of whole genome

First, we filter the GFF file to keep only genes.
```{r filter gff genome}
genome_gff <- gff[
  gff$type == "gene", 
]
```

We can check the source of the filtered annotations:
```{r gff_genome table of annotation source}
table(genome_gff$source)
```

And the number of genes:

```{r gff_genome number of genes}
nrow(genome_gff)
length(unique(genome_gff$ID))
```

We order the entries:

```{r order genome_gff}
genome_gff <- genome_gff[order(
  genome_gff$seqnames,                     # First, order by chromosome
  genome_gff$strand,                       # Then, order by strand
  ifelse(genome_gff$strand == "+",         # If strand is "+", order by start
         genome_gff$start, 
         -genome_gff$end)                  # If strand is "-", order by end in descending order
), ]

# View the first few rows of the ordered dataframe by strand
head(genome_gff[genome_gff$strand == "+",])
head(genome_gff[genome_gff$strand == "-",])
```

Now, we can filter to keep only genes that are not overlapping with each other. We first look for genes that overlap:

```{r}
# Initialize a vector to store overlap results
genome_gff$overlap <- FALSE

# Loop over genes, comparing each gene with the next one in the same strand
for (i in 1:(nrow(genome_gff) - 1)) {

  # Check if genes are on the same chromosome/scaffold and same strand
  if (genome_gff$seqnames[i] == genome_gff$seqnames[i + 1] &&
      genome_gff$strand[i] == genome_gff$strand[i + 1]){

    # Check overlap if strand is positive
    if (genome_gff$strand[i] == "+" &&
        genome_gff$end[i] >= genome_gff$start[i+1]) {
      genome_gff$overlap[i] <- TRUE
      genome_gff$overlap[i+1] <- TRUE
    }

     # Check overlap if strand is negative
    if (genome_gff$strand[i] == "-" &&
       genome_gff$start[i] <= genome_gff$end[i+1]) {
      genome_gff$overlap[i] <- TRUE
      genome_gff$overlap[i+1] <- TRUE
    }
  }

}

# Number of overlapping genes
nrow(genome_gff[genome_gff$overlap == TRUE,])
```

We remove the overlapping genes from `genome_gff` and check the number of remaining genes:

```{r gff_genome remove overlap genes}
# Remove overlapping genes from genome_gff
genome_gff <- genome_gff[!genome_gff$overlap, ]
nrow(genome_gff)
length(unique(genome_gff$ID))
```

### Chromosome lengths

To define the upstream regions the chromosome lengths are needed to be the maximum for the start of the upstream region of the genes on the negative strand.

```{r load gbff}
gbff_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/ncbi_dataset/ncbi_dataset/data/GCF_000238955.4/genomic.gbff" 
file_lines <- readLines(gbff_file)
# Extract the LOCUS line
locus_lines <- file_lines[grep("^LOCUS", file_lines)]
head(locus_lines)
```

```{r explore genome_gff versions}
length(unique(genome_gff$seqnames))
# Remove version suffix from seqnames
length(unique(sub("\\.\\d+$", "", genome_gff$seqnames)))
```

It does not have several versions, so we can remove the version numbers from `seqnames` in the `genome_gff` and from variable `dpp4_chr`.

```{r remove version from seqnames}
# Remove version numbers from seqnames in the filtered files
genome_gff$seqnames <- sub("\\.\\d+$", "", genome_gff$seqnames)
dpp4_chr <- "NC_036801"
```

```{r extract chrom names and lengths}
# Extract chromosome names and lengths
chrom_info <- gsub("^LOCUS\\s+(\\S+)\\s+(\\d+)\\s+.*$", "\\1,\\2", locus_lines)
rm(locus_lines)
chrom_info <- strsplit(chrom_info, ",")
chrom_names <- sapply(chrom_info, `[`, 1)
chrom_lengths <- as.numeric(sapply(chrom_info, `[`, 2))
# Create a named vector
chrom_lengths <- setNames(chrom_lengths, chrom_names)
```

```{r check correspondence genome_gff and chrom_lengths}
# Check how many chromosomes are in chrom_names but not in genome_gff$seqnames
length(setdiff(chrom_names, unique(genome_gff$seqnames)))

# You can also check the reverse - how many chromosomes in genome_gff$seqnames are not in chrom_names
length(setdiff(unique(genome_gff$seqnames), chrom_names))
```

### Define the 200 kb upstream genes of whole genome

```{r}
# Initialize the new columns for upstream start and end in genome_gff
genome_gff$upstream_start <- NA
genome_gff$upstream_end <- NA
genome_gff$upstream_length <- NA

# Process each chromosome separately
for (chr in unique(genome_gff$seqnames)) {
  chr_length <- chrom_lengths[chr]
  chr_genes <- genome_gff[genome_gff$seqnames == chr, ]

  ### Process positive strand genes ###

  pos_genes <- chr_genes[chr_genes$strand == "+", ]
  if (nrow(pos_genes) > 0) {

    # First gene in chromosome (positive strand)
    if (pos_genes$start[1] > 1){
      pos_genes$upstream_end[1] <- pos_genes$start[1] - 1
      pos_genes$upstream_start[1] <- max(1, pos_genes$start[1] - 20000)
      pos_genes$upstream_length[1] <- pos_genes$upstream_end[1] - pos_genes$upstream_start[1] + 1

      # Only keep upstream region if it's valid (at least 1 base)
      if (pos_genes$upstream_end[1] - pos_genes$upstream_start[1] < 0){
        pos_genes$upstream_start[1] <- NA
        pos_genes$upstream_end[1] <- NA
        pos_genes$upstream_length[1] <- NA
      }
    }

    # Remaining genes
    if (nrow(pos_genes) > 1) {
      for (i in 2:nrow(pos_genes)) {
        pos_genes$upstream_end[i] <- pos_genes$start[i] - 1
        pos_genes$upstream_start[i] <- max(pos_genes$start[i] - 20000, pos_genes$end[i-1] + 1)
        pos_genes$upstream_length[i] <- pos_genes$upstream_end[i] - pos_genes$upstream_start[i] + 1

        # Only keep upstream region if it's valid (at least 1 base)
        if (pos_genes$upstream_end[i] - pos_genes$upstream_start[i] < 0){
          pos_genes$upstream_start[i] <- NA
          pos_genes$upstream_end[i] <- NA
          pos_genes$upstream_length[i] <- NA
        }
      }
    }
  }

  ### Process negative strand genes ###
  neg_genes <- chr_genes[chr_genes$strand == "-", ]
  if (nrow(neg_genes) > 0) {

    # First gene in chromosome (negative strand)
    if (neg_genes$end[1] < chr_length) {
      neg_genes$upstream_start[1] <- min(chr_length, neg_genes$end[1] + 20000)
      neg_genes$upstream_end[1] <- neg_genes$end[1] + 1
      neg_genes$upstream_length[1] <- neg_genes$upstream_start[1] - neg_genes$upstream_end[1] + 1

      # Only keep upstream region if it's valid (at least 1 base)
      if (neg_genes$upstream_start[1] - neg_genes$upstream_end[1] < 0) {
        neg_genes$upstream_start[1] <- NA
        neg_genes$upstream_end[1] <- NA
        neg_genes$upstream_length[1] <- NA
      }
    }

    # Remaining genes
    if (nrow(neg_genes) > 1) {
      for (i in 2:nrow(neg_genes)) {
        neg_genes$upstream_start[i] <- min(neg_genes$end[i] + 20000, neg_genes$start[i-1] - 1)
        neg_genes$upstream_end[i] <- neg_genes$end[i] + 1
        neg_genes$upstream_length[i] <- neg_genes$upstream_start[i] - neg_genes$upstream_end[i] + 1

        # Only keep upstream region if it's valid (at least 1 base)
        if (neg_genes$upstream_start[i] - neg_genes$upstream_end[i] < 0) {
          neg_genes$upstream_start[i] <- NA
          neg_genes$upstream_end[i] <- NA
          neg_genes$upstream_length[i] <- NA
        }
      }
    }
  }

  # Update the main genome_gff dataframe
  if(nrow(pos_genes)>0){
    genome_gff[genome_gff$seqnames == chr & genome_gff$strand == "+", c("upstream_start", "upstream_end", "upstream_length")] <- 
    pos_genes[, c("upstream_start", "upstream_end", "upstream_length")]
  }

  if(nrow(neg_genes)>0){
    genome_gff[genome_gff$seqnames == chr & genome_gff$strand == "-", c("upstream_start", "upstream_end", "upstream_length")] <- 
    neg_genes[, c("upstream_start", "upstream_end", "upstream_length")]
  }

}

# Check the results
summary(genome_gff$upstream_length)
```

#### Subset chromosome of interest

```{r filter gff chromosome alternative}
chromosome_gff <- genome_gff[
  genome_gff$seqnames == dpp4_chr,
]
```

We can check the source of the filtered annotations:
```{r gff_chromosome table of annotation source}
table(chromosome_gff$source)
```

Check the number of remaining genes:

```{r gff_chromosome remaining genes}
nrow(chromosome_gff)
length(unique(chromosome_gff$ID))
```

Check the summary of the upstream lengths:

```{r}
summary(chromosome_gff$upstream_length)
```

#### Subset region of interest

```{r filter gff region alternative}
region_gff <- chromosome_gff[
  chromosome_gff$start <= region_end & 
  chromosome_gff$end >= region_start, 
]
```

We can check the source of the filtered annotations:
```{r gff_region table of annotation source}
table(region_gff$source)
```

Check the number of remaining genes:

```{r gff_region remaining genes}
nrow(region_gff)
length(unique(region_gff$ID))
```

We can visualize the genes on the region of interest:

```{r gff_region plot}
# Extract relevant columns
gene_data <- region_gff[, c("strand","start", "end", "gene")]

@@

# Set up larger plotting area and adjust margins
par(mar=c(6, 10, 4, 2), xpd=TRUE)

@@

# Add a legend at the top-left corner
legend("topright", 
       legend=c("Positive strand (blue)", "Negative strand (red)"), 
       fill=c("blue", "red"), 
       bty="n", cex=0.8)
```

Check the summary of the upstream lengths:

```{r summary of region upstream length}
summary(region_gff$upstream_length)
```

@@

```{r load cgi annotation file}
# Load the CGI annotation file
cgi_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/CGI-Mzebra.txt" 
cgif <- read.table(cgi_file, header = TRUE, sep = "\t")
head(cgif)
```
Lets remove the chromosome version number from the CGI annotation:

```{r}
cgif$chr <- sub("\\.\\d+$", "", cgif$chr)
```

Check the summary of the island lengths:

```{r summary of cgi lengths}
summary(cgif$length)
```

Not all chromosomes in `genome_gff` are in the CGI annotations:

```{r}
# Identify seqnames in genome_gff that are not in cgif
missing_seqnames <- setdiff(unique(genome_gff$seqnames), unique(cgif$chr))

# Print the missing seqnames
print(missing_seqnames)
```

To check when extracting chromosome info.
```{r}
nrow(cgif[cgif$chr==missing_seqnames[1],])
```

Lets visualize the islands in the genomic region of interest:

@@

For each position within each upstream region we check whether it is in a CGI and the distance to the gene.

```{r, eval = FALSE}
# Function to efficiently check if positions are within any CGI
is_within_cgi_vector <- function(positions, cgi_intervals) {
  # Apply the check for each position against all intervals
  sapply(positions, function(pos) {
    any(pos >= cgi_intervals[,1] & pos <= cgi_intervals[,2])
  })
}

# Initialize results list
#results_list <- vector("list", nrow(region_gff))
## TODO: Update comment
cgiVsDistance <- matrix(data = NA, nrow = nrow(genome_gff), ncol = 20000)

# Get the genome_gff indices for genes with a defined upstream region
nonNA_upstream <- which(!is.na(genome_gff$upstream_length))

# Set start variable for previous chromosome name
prev_chrom_name <- 0

for (i in nonNA_upstream){
  # Get current chromosome name
  chrom_name <- genome_gff$seqnames[i]
  # Get the cgi annotations for the current chromosome
  if (prev_chrom_name != chrom_name){
    chrom_cgi <- cgif[cgif$chr==chrom_name,]
    # Create a matrix of CGI intervals for faster lookup
    chrom_cgi_intervals <- cbind(chrom_cgi$start, chrom_cgi$end)
  }

  # Get the sequence of upstream positions
  upstream_start <- genome_gff$upstream_start[i]
  upstream_end <- genome_gff$upstream_end[i]
  if (upstream_start > upstream_end) { # Upstream in the negative strand
    positions <- seq(from = upstream_end, to = upstream_start, by = 1)
  } else { # Upstream in the positive strand
    positions <- seq(from = upstream_start, to = upstream_end, by = 1)
  }

  # Get CGI status for all positions at once
  in_cgi <- is_within_cgi_vector(positions, chrom_cgi_intervals)

  # Get the sequence of upstream distances to the gene start
  if(genome_gff$strand[i] == "+") {
    # For positive strand: distance from gene start
    distances <- genome_gff$start[i] - positions
  } else {
    # For negative strand: distance from gene end
    distances <- positions - genome_gff$end[i]
  }

  # Save the CGI status in the distance matrix
  cgiVsDistance[i,distances] <- in_cgi

  # Update the variable for the previous chromosome name
  prev_chrom_name <- chrom_name
}
```

```{r, eval=FALSE}
save(cgiVsDistance, file = "cgiVsDistance.RData")
```

```{r}
load("cgiVsDistance.RData")
```

Now we can calculate the probability of each promoter site (described by the distance to the gene) to be in a CGI:

```{r}
cgiProbByDistance <- colMeans(cgiVsDistance, na.rm = TRUE)
distCounts <- colSums(!is.na(cgiVsDistance))
```

```{r, fig.height=12, fig.width=10}
# Set up the plot with better formatting
par(mfrow = c(2,1))  # Set up 2 plots vertically
par(mar = c(4, 5, 4, 2))  # Adjust margins for top plot
# Plot 1: Probability plot
plot(1:length(cgiProbByDistance), cgiProbByDistance, 
     xlab = "Distance of Upstream Positions to Gene",  
     ylab = "Probability of Position Being in CGI",
     main = "Probability of Position Being in CpG Island\nby Distance to Gene Start",
     cex.lab = 1.2,      # Larger axis labels
     cex.axis = 1.1,     # Larger axis numbers
     cex.main = 1.3,     # Larger title
     las = 1,            # Horizontal axis labels
     pch = 16,          # Solid circle points
     ylim = c(0, 1),    # Force probability y-axis to [0,1]
     col = adjustcolor("darkblue", alpha.f = 0.5))  # Semi-transparent blue

# Plot 2: Count plot
par(mar = c(5, 5, 3, 2))  # Adjust margins for bottom plot
plot(1:length(distCounts), distCounts, type = "h",  # Using the correct counts
     xlab = "Distance to Gene Start (bp)",
     ylab = "Number of Positions",
     main = "Number of Positions at Each Distance",
     col = "darkgreen",
     lwd = 1,
     ylim = c(0, max(distCounts) * 1.05))  # Force y-axis to start at 0 with 5% padding
grid(col = "gray90", lty = "dotted")
```

#### Subset chromosome of interest

```{r}
cgiProbByDistance <- colMeans(cgiVsDistance[which(genome_gff$seqnames == dpp4_chr),], na.rm = TRUE)
distCounts <- colSums(!is.na(cgiVsDistance[which(genome_gff$seqnames == dpp4_chr),]))
```

```{r, fig.height=12, fig.width=10}
# Set up the plot with better formatting
par(mfrow = c(2,1))  # Set up 2 plots vertically
par(mar = c(4, 5, 4, 2))  # Adjust margins for top plot
# Plot 1: Probability plot
plot(1:length(cgiProbByDistance), cgiProbByDistance, 
     xlab = "Distance of Upstream Positions to Gene",  
     ylab = "Probability of Position Being in CGI",
     main = "Probability of Position Being in CpG Island\nby Distance to Gene Start",
     cex.lab = 1.2,      # Larger axis labels
     cex.axis = 1.1,     # Larger axis numbers
     cex.main = 1.3,     # Larger title
     las = 1,            # Horizontal axis labels
     pch = 16,          # Solid circle points
     ylim = c(0, 1),    # Force probability y-axis to [0,1]
     col = adjustcolor("darkblue", alpha.f = 0.5))  # Semi-transparent blue

# Plot 2: Count plot
par(mar = c(5, 5, 3, 2))  # Adjust margins for bottom plot
plot(1:length(distCounts), distCounts, type = "h",  # Using the correct counts
     xlab = "Distance to Gene Start (bp)",
     ylab = "Number of Positions",
     main = "Number of Positions at Each Distance",
     col = "darkgreen",
     lwd = 1,
     ylim = c(0, max(distCounts) * 1.05))  # Force y-axis to start at 0 with 5% padding
grid(col = "gray90", lty = "dotted")
```

#### Subset region of interest

```{r}
region_indeces <- genome_gff$seqnames == dpp4_chr & genome_gff$start <= region_end & genome_gff$end >= region_start
cgiProbByDistance <- colMeans(cgiVsDistance[region_indeces,], na.rm = TRUE)
distCounts <- colSums(!is.na(cgiVsDistance[region_indeces,]))
```

```{r, fig.height=12, fig.width=10}
# Set up the plot with better formatting
par(mfrow = c(2,1))  # Set up 2 plots vertically
par(mar = c(4, 5, 4, 2))  # Adjust margins for top plot
# Plot 1: Probability plot
plot(1:length(cgiProbByDistance), cgiProbByDistance, 
     xlab = "Distance of Upstream Positions to Gene",  
     ylab = "Probability of Position Being in CGI",
     main = "Probability of Position Being in CpG Island\nby Distance to Gene Start",
     cex.lab = 1.2,      # Larger axis labels
     cex.axis = 1.1,     # Larger axis numbers
     cex.main = 1.3,     # Larger title
     las = 1,            # Horizontal axis labels
     pch = 16,          # Solid circle points
     ylim = c(0, 1),    # Force probability y-axis to [0,1]
     col = adjustcolor("darkblue", alpha.f = 0.5))  # Semi-transparent blue

# Plot 2: Count plot
par(mar = c(5, 5, 3, 2))  # Adjust margins for bottom plot
plot(1:length(distCounts), distCounts, type = "h",  # Using the correct counts
     xlab = "Distance to Gene Start (bp)",
     ylab = "Number of Positions",
     main = "Number of Positions at Each Distance",
     col = "darkgreen",
     lwd = 1,
     ylim = c(0, max(distCounts) * 1.05))  # Force y-axis to start at 0 with 5% padding
grid(col = "gray90", lty = "dotted")
```
